Prospects of detecting gamma-ray emission from galaxy clusters: cosmic rays and dark matter annihilations

### 2011-06-22 paper.tex
big footnote moved to text

\subsubsection{Substructures}
\label{sect:subst}
High-resolution dissipationless DM simulations of MW type halos find substantial
amount of substructures in the periphery of DM halos, while the substructures in
the center suffer from dynamical friction and tidal effects depleting their
central number densities. Since the DM annihilation rate depends on the density
squared, the resulting flux from substructures is boosted compared to the smooth
density distribution. Recent high resolution simulations suggest a flux
enhancement from substructures of more than 100 \cite{2008MNRAS.391.1685S,
  2010ApJ...718..899A} for a MW size halo. We adopt a boost from substructures
of 230 which is the enhancement found in one of the highest resolved MW halos
ever simulated.\footnote{\C{While there is still a discrepancy in the literature
    of the exact value of the predicted boost factor of the DM annihilation
    luminosity from substructures in DM halos, this inconsistency starts to
    become resolved, apparently converging toward predictions at the high end,
    e.g. those by the Aquarius simulations \protect \cite{2008MNRAS.391.1685S,
      2008Natur.456...73S}. The initial suggestions of a small {\em total} boost
    ($<R_{200}$) of order unity by the Via Lactea simulations
    \cite{2008JPhCS.125a2008K} made assumptions of computing the boost assuming
    that the substructure luminosity follows that of the smooth DM
    distribution. Physically, this would imply substructure to follow the smooth
    DM distribution and a radially independent concentration parameter of
    substructures; both following from neglecting tidal mass loss of
    satellites. Tidal losses, however, are at work in simulations and cause the
    radial number density to be anti-biased with respect to the host's mass
    density profile yielding a significantly flatter subhalo distribution
    compared to the smooth DM distribution \cite{2010ApJ...718..899A}. This
    increases the substructure boost preferentially in the outer parts of DM
    halos with a factor that ranges between 20 and 1000 (for dwarfs and galaxy
    clusters), hence depending sensitively on the host halo mass (see
    Figure~\ref{fig:radial_lum}).}}  We use a double power-law function to fit
the luminosity from the smooth component of substructures (i.e. substructures
within substructures are not included) inside radius $r$ which is determined for
the Aquarius simulations
\cite{2008MNRAS.391.1685S,2008Natur.456...73S}.\footnote{Our approach of fitting
  the scaling behavior of $L_\sub(<r)$ directly from numerical simulations
  self-consistently accounts for the radial dependence of the substructure
  concentration \C{due to tidal mass losses} \protect
  \cite{2008MNRAS.391.1685S}.} Our best fit is given by
  L_\sub(<x) &=& a_0\,C(\mvir)\,L_{200\sm}(\mvir)\,x^{a_1
    \displaystyle{x}^{a_2}}\,,\label{eq:Lsub}\\
   a_0 &=& 0.76\,,\quad a_1 = 0.95\,,\quad  a_2 = -0.27\,, \quad 
 \nonumber\\
  C(\mvir) &=& \left(\frac{\mvir\,M_{\rm res,sim}}
{M_{\rm 200sim}\,M_{\rm lim}}\right)^{\alpha_C}
=0.023\,\left(\frac{\mvir}{M_{\rm lim}}\right)^{\alpha_C}.\nonumber\\
\label{eq:Csub}
Here, $\alpha_C=0.226$, $a_i$ denote our fit variables, $L_{200\sm}$
is the luminosity from the smooth halo without substructures within
$\rvir$ and $x= r/\rvir$.  We derive the normalization function
$C(\mvir)$ in Eq.~(\ref{eq:Csub}) from the simulations in reference
\cite{2008Natur.456...73S} using a value of $M_{\rm
  200sim}=1.9\times10^{12}\msun$ for the mass of the MW halo in the
simulation and $M_{\rm res,sim}=10^5\msun$ for the mass of the
smallest resolved subhalos in the MW simulation.  The smallest mass of
subhalos in reality, $M_{\rm lim}$, is determined from the free
streaming length of DM at decoupling---an effect that erases structure
on scales smaller than the free streaming length. In applying
Eq.~(\ref{eq:Csub}), we implicitly assume that the mass power-law
scaling relation is valid down to the free streaming mass of DM
halos. In the CDM universe, this is conventionally taken to be
$10^{-6}\msun$ \cite{2001PhRvD..64h3507H, 2005JCAP...08..003G} (see
\cite{2009NJPh...11j5027B} for a discussion of the range expected in
various DM models). Note that potentially the power-law could flatten
towards smaller mass scales although current simulations show no hints
of such a behavior and since Einstein's gravity is a scale-free
theory, we do not expect such a behavior on theoretical grounds
either.  For DM halos more (less) massive than the MW we expect a
larger (smaller) boost from substructures, simply because of the
larger (smaller) mass range down to the minimum mass $M_\rmn{lim}$.

>>>>

\subsubsection{Substructures}
\label{sect:subst}
High-resolution dissipationless DM simulations of MW type halos find substantial
amount of substructures in the periphery of DM halos, while the substructures in
the center suffer from dynamical friction and tidal effects depleting their
central number densities. Since the DM annihilation rate depends on the density
squared, the resulting flux from substructures is boosted compared to the smooth
density distribution. Recent high resolution simulations suggest a flux
enhancement from substructures of more than 100 \cite{2008MNRAS.391.1685S,
  2010ApJ...718..899A} for a MW size halo. We adopt a boost from substructures
of 230 which is the enhancement found in one of the highest resolved MW halos
ever simulated. \C{While there is still a discrepancy in the literature of the
  exact value of the predicted boost factor of the DM annihilation luminosity
  from substructures in DM halos, this inconsistency starts to become resolved,
  apparently converging toward predictions at the high end, e.g. those by the
  Aquarius simulations \protect \cite{2008MNRAS.391.1685S,
    2008Natur.456...73S}. The initial suggestions of a small {\em total} boost
  ($<R_{200}$) of order unity by the Via Lactea simulations
  \cite{2008JPhCS.125a2008K} made assumptions of computing the boost assuming
  that the substructure luminosity follows that of the smooth DM
  distribution. Physically, this would imply substructure to follow the smooth
  DM distribution and a radially independent concentration parameter of
  substructures; both following from neglecting tidal mass loss of
  satellites. Tidal losses, however, are at work in simulations and cause the
  radial number density to be anti-biased with respect to the host's mass
  density profile yielding a significantly flatter subhalo distribution compared
  to the smooth DM distribution \cite{2010ApJ...718..899A}. This increases the
  substructure boost preferentially in the outer parts of DM halos with a factor
  that ranges between 20 and 1000 (for dwarfs and galaxy clusters), hence
  depending sensitively on the host halo mass (see
  Figure~\ref{fig:radial_lum}).}

We use a double power-law function to fit the luminosity from the smooth
component of substructures (i.e. substructures within substructures are not
included) inside radius $r$ which is determined for the Aquarius simulations
\cite{2008MNRAS.391.1685S,2008Natur.456...73S}.\footnote{Our approach of fitting
  the scaling behavior of $L_\sub(<r)$ directly from numerical simulations
  self-consistently accounts for the radial dependence of the substructure
  concentration \C{due to tidal mass losses} \protect
  \cite{2008MNRAS.391.1685S}.} Our best fit is given by
\begin{eqnarray}
  L_\sub(<x) &=& a_0\,C(\mvir)\,L_{200\sm}(\mvir)\,x^{a_1
    \displaystyle{x}^{a_2}}\,,\label{eq:Lsub}\\
   a_0 &=& 0.76\,,\quad a_1 = 0.95\,,\quad  a_2 = -0.27\,, \quad 
 \nonumber\\
  C(\mvir) &=& \left(\frac{\mvir\,M_{\rm res,sim}}
{M_{\rm 200sim}\,M_{\rm lim}}\right)^{\alpha_C}
=0.023\,\left(\frac{\mvir}{M_{\rm lim}}\right)^{\alpha_C}.\nonumber\\
\label{eq:Csub}
\end{eqnarray}
Here, $\alpha_C=0.226$, $a_i$ denote our fit variables, $L_{200\sm}$
is the luminosity from the smooth halo without substructures within
$\rvir$ and $x= r/\rvir$.  We derive the normalization function
$C(\mvir)$ in Eq.~(\ref{eq:Csub}) from the simulations in reference
\cite{2008Natur.456...73S} using a value of $M_{\rm
  200sim}=1.9\times10^{12}\msun$ for the mass of the MW halo in the
simulation and $M_{\rm res,sim}=10^5\msun$ for the mass of the
smallest resolved subhalos in the MW simulation.  The smallest mass of
subhalos in reality, $M_{\rm lim}$, is determined from the free
streaming length of DM at decoupling---an effect that erases structure
on scales smaller than the free streaming length. In applying
Eq.~(\ref{eq:Csub}), we implicitly assume that the mass power-law
scaling relation is valid down to the free streaming mass of DM
halos. In the CDM universe, this is conventionally taken to be
$10^{-6}\msun$ \cite{2001PhRvD..64h3507H, 2005JCAP...08..003G} (see
\cite{2009NJPh...11j5027B} for a discussion of the range expected in
various DM models). Note that potentially the power-law could flatten
towards smaller mass scales although current simulations show no hints
of such a behavior and since Einstein's gravity is a scale-free
theory, we do not expect such a behavior on theoretical grounds
either.  For DM halos more (less) massive than the MW we expect a
larger (smaller) boost from substructures, simply because of the
larger (smaller) mass range down to the minimum mass $M_\rmn{lim}$.
